---
title: "Does a Stimulant Increase Locomotion Within the Same Mice? A Paired-Samples t-Test"
subtitle: "Lab notebook template: paired-samples (dependent) t-test, with a lesson on many measurements per animal"
author: "Your Name"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: show
    code-tools: true
    embed-resources: true
    smooth-scroll: true
    fig-width: 6.5
    fig-height: 4.2
    fig-dpi: 300
    fig-align: center
    tbl-cap-location: top
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note icon=false}
## How to use this notebook

This is a **template lab notebook** that walks through a complete analysis: a question, a plan, the data, the code, and a written report of the results. To make it yours:

1. Open `paired_t_test.Rproj` first (so RStudio uses this folder as the working directory).
2. Replace `Your Name` in the header above with your name.
3. Click **Render** to build the notebook. Every number, table, and figure is regenerated from the data and code.
4. Read the text, then change things and re-render to see what happens (see @sec-your-turn).

The data are **simulated** for teaching purposes. No animals were tested. A special-topic section (@sec-nested) is a short lesson on a very common mistake: treating several measurements from the *same animal* as if they came from different animals.
:::

| Notebook details | |
|:--|:--|
| **Project** | Paired-samples (dependent) t-test |
| **Data files** | `Data/open_field_crossover.csv` (main analysis) and `Data/slice_recordings.csv` (nested-data lesson) |
| **Analysis script** | `R/02_analysis.R` (same analysis, with extra teaching comments) |
| **Helper code** | `R/00_setup.R` (packages, plot theme, formatting helpers) |
| **Software** | `r R.version.string`; Quarto |

```{r}
#| label: setup
# Load packages, the plot theme, and helper functions (see R/00_setup.R)
source("R/00_setup.R")
```

::: {.callout-note collapse="true"}
## A note on how these data were chosen
Both data sets were generated by one script (`R/01_simulate_data.R`) with a single random seed. The seed was chosen by a rule fixed in advance: the first seed (trying 1, 2, 3, ...) for which the mean of the analyzed paired differences (open field) and the mean neuron-level change (slices) were each within 10% of the true effect built into the simulation, and the analyzed differences did not look non-normal. The rule keeps the examples typical, but it is still a choice, and a real study does not get to make it. A different seed would give somewhat different numbers, and with samples this small the *p*-values can change a lot. The sample sizes (14 mice; 8 mice with 34 neurons) were also chosen to suit a classroom example, not planned with a power analysis (see @sec-power-plan).
:::

# Problem Statement and Hypotheses {#sec-problem}

## Background

In an **open-field test**, a mouse is placed in an open arena and its movement is tracked. The total **distance traveled** is a common measure of general locomotor activity. Stimulant drugs are expected to increase it.

Mice differ enormously in their baseline activity: some are naturally restless and others are calm. If we gave the drug to one group of mice and a saline injection (the "vehicle") to a different group, those natural differences between animals would add noise to the comparison. A **within-subject** (repeated-measures) design avoids much of that noise by testing **each mouse under both conditions**, so each animal serves as its own control.

In this notebook, 14 mice were each tested twice, one week apart: once after a low dose of a stimulant (**Drug**) and once after a saline injection (**Vehicle**). Half of the mice received the drug first and half received the vehicle first. This is called a **counterbalanced crossover design**, and the counterbalancing keeps the order of the sessions from being confused with the effect of the drug (see @sec-order).

## Problem statement

> **Do mice travel a different distance in the open field after receiving the drug than after receiving the vehicle?**

Each mouse contributes **two related measurements**, so the observations are **not independent**: the two distances from the same mouse tend to be similar. The appropriate test is a **paired-samples t-test** (also called a dependent-samples or repeated-measures t-test).

## Hypotheses

The paired t-test works on each mouse's **difference score**, $D = \text{Drug} - \text{Vehicle}$, and asks whether the population mean of those differences, $\mu_D$, is zero.

$$
H_0: \mu_D = 0 \qquad \text{(on average, the drug does not change distance traveled)}
$$

$$
H_1: \mu_D \neq 0 \qquad \text{(on average, the drug changes distance traveled)}
$$

Here $\mu_D$ is the **population** mean difference (what we would see if we could test every mouse of this type under both conditions). We estimate it from our sample. Because the difference is defined as **Drug minus Vehicle**, a positive difference means the mouse traveled *farther* after the drug.

We use a **two-tailed** test because a drug could, in principle, change activity in either direction, and we set the **significance level** at $\alpha = .05$.

::: {.callout-tip}
## A paired t-test is a one-sample t-test on the differences
If you remember the one-sample t-test from earlier in the course, this is the same test. We compute one difference per mouse and test whether the mean of those differences is zero. The "pairing" is handled by the design of the data, not by new math.
:::

## Decision errors: Type I and Type II {#sec-errors}

Any decision based on samples can be wrong. @tbl-errors shows the four possible outcomes of a hypothesis test.

```{r}
#| label: tbl-errors
#| tbl-cap: "Possible outcomes of a hypothesis test"
tribble(
  ~Decision,                 ~`H0 is actually true`,                         ~`H0 is actually false`,
  "Reject H0",               "Type I error (probability = α)",                "Correct decision (probability = power = 1 − β)",
  "Fail to reject H0",       "Correct decision (probability = 1 − α)",        "Type II error (probability = β)"
) |>
  pub_table(align = "lcc")
```

- A **Type I error** (false positive) means concluding that the drug changes activity when, in the population, it does not. We control this rate by choosing $\alpha$.
- A **Type II error** (false negative) means missing a real effect of the drug. Its probability is $\beta$, and **power** $= 1 - \beta$ is the probability of detecting a real effect of a given size. We study both in @sec-sim.

# Methods {#sec-methods}

## Subjects and design

Fourteen adult mice were tested in a **counterbalanced crossover design**. Each mouse completed two open-field sessions one week apart (the gap, called a *washout period*, lets the drug clear before the second session). Before one session a mouse received an injection of a low dose of a stimulant (**Drug**); before the other it received a saline injection (**Vehicle**). Mice were randomly assigned to one of two sequences: **Drug first** or **Vehicle first**, with equal numbers planned in each. *(Simulated data for teaching. In a real notebook, give the species/strain, sex, age, source, housing, the drug, dose, route and timing of injection, how animals were randomized, and the approval number of your animal-care protocol.)*

## Procedure

Each mouse was placed in the center of the arena and allowed to explore for 10 minutes. Video tracking recorded the total distance traveled in meters. *(In a real notebook, include the arena size, lighting, time of day, and whether the scorer was blind to the treatment. Blind scoring protects against unintentional bias.)*

## Outcome measure

The outcome was the **total distance traveled (m)** in the 10-minute session (higher values mean more locomotion).

## Missing data

For one mouse (`M07`), the video tracking failed during its second session, so that distance is recorded as missing (`NA`). A paired analysis needs **both** measurements from a mouse, so that mouse has to be left out of the paired test entirely, even though its other session is fine. We report this loss. Missing values are never recorded as 0, because that would claim the mouse did not move at all.

## Data description {#sec-data}

The raw data are stored in `Data/open_field_crossover.csv`. The file is in **tidy "long" format**: one row per mouse per session. The raw file is **never edited by hand**: all cleaning and calculations are done in code so they are documented and repeatable.

```{r}
#| label: tbl-dictionary
#| tbl-cap: "Variables in the raw data file"
tribble(
  ~Variable,         ~Type,          ~Description,
  "mouse_id",        "Text",         "Unique animal identifier (each mouse appears in two rows)",
  "sex",             "Category",     "F = female, M = male",
  "sequence",        "Category",     "Order of treatments: Drug first or Vehicle first",
  "session",         "Whole number", "1 = first session, 2 = second session (one week later)",
  "treatment",       "Category",     "Treatment given before that session: Drug or Vehicle",
  "distance_m",      "Number",       "Distance traveled in the 10-minute session, in meters (NA = tracking failed)"
) |>
  pub_table(align = "lll")
```

We import the data and immediately look at it: its structure, its ranges, and whether anything is missing or duplicated.

```{r}
#| label: import
alpha        <- 0.05   # significance level (accepted Type I error rate)
target_power <- 0.80   # conventional minimum power

of_raw <- read_csv(here("Data", "open_field_crossover.csv"),
                   show_col_types = FALSE)

glimpse(of_raw)
```

```{r}
#| label: checks
# Each mouse should appear exactly twice, once under each treatment.
tibble(
  `Rows`                         = nrow(of_raw),
  `Mice`                         = n_distinct(of_raw$mouse_id),
  `Missing distances`            = sum(is.na(of_raw$distance_m)),
  `Mice not seen exactly twice`  = sum(count(of_raw, mouse_id)$n != 2),
  `Mice missing a treatment`     = sum(of_raw |> group_by(mouse_id) |>
                                         summarize(k = n_distinct(treatment)) |> pull(k) != 2)
) |>
  pub_table(align = "ccccc")
```

For a paired analysis we need **one row per mouse** with the two measurements side by side. The function `pivot_wider()` reshapes the long data into that "wide" layout, and we then compute each mouse's difference. Mice with a missing value are dropped at this step.

```{r}
#| label: prepare
of <- of_raw |>
  mutate(
    treatment = factor(treatment, levels = c("Vehicle", "Drug")),
    sex       = factor(sex, levels = c("F", "M"), labels = c("Female", "Male"))
  )

# One row per mouse: Vehicle and Drug distances side by side
of_wide <- of |>
  select(mouse_id, sex, sequence, treatment, distance_m) |>
  pivot_wider(names_from = treatment, values_from = distance_m) |>
  mutate(diff = Drug - Vehicle)           # Drug minus Vehicle

lost_mice <- of_wide |> filter(is.na(diff))
pairs     <- of_wide |> drop_na(diff)     # complete pairs only

n_recorded <- nrow(of_wide)
n_pairs    <- nrow(pairs)
lost_row   <- of |> filter(is.na(distance_m))   # which session was lost?
```

Of the `r n_recorded` mice tested, `r nrow(lost_mice)` (`r paste(lost_mice$mouse_id, collapse = ", ")`) lost its **`r tolower(as.character(lost_row$treatment))`** session recording, leaving **`r n_pairs` complete pairs** for the analysis. The number of pairs, not the number of measurements, is the sample size for a paired test.

# Statistical Analysis {#sec-stats}

## Planned analysis

- **Test.** A two-tailed **paired-samples t-test**, with $\alpha = .05$. It is a one-sample t-test on the difference scores:
$$t = \frac{\bar{D}}{s_D / \sqrt{n}}, \qquad df = n - 1,$$
where $\bar{D}$ is the mean of the differences, $s_D$ is their standard deviation, and $n$ is the number of **pairs**.
- **Assumptions.** (1) The pairs are independent of one another: different mice do not influence each other, and each mouse supplies exactly one pair; (2) the outcome is continuous; (3) the **difference scores** are approximately normal (checked with a histogram, a Q-Q plot, and the Shapiro–Wilk test); (4) no extreme outliers among the differences. Notice that the assumption of normality applies to the *differences*, not to the two sets of distances separately, and that there is no "equal variances" assumption.
- **Effect size.** The **raw mean difference** (in meters), and **Cohen's $d_z$** with a 95% confidence interval, $d_z = \bar{D}/s_D$ (the mean difference in units of the standard deviation of the differences). See @sec-effect for how $d_z$ differs from the *d* in the two-sample project.
- **Confidence interval.** A 95% confidence interval for the mean difference.
- **Robustness check.** A Wilcoxon signed-rank test, a rank-based test that does not assume the differences are normal. (It does assume the differences are roughly symmetric around their center, and it tests whether they tend to be shifted away from zero rather than testing the mean.)
- **Software.** Analyses were run in `r R.version.string` using the tidyverse, `here`, `broom`, `rstatix`, `effectsize`, `pwr`, `knitr`, `kableExtra`, and `patchwork`.

## Power and sample size plan {#sec-power-plan}

**Power** is the probability of rejecting $H_0$ when it is false. It depends on the effect size, the number of pairs, $\alpha$, and whether the test is one- or two-tailed. Good practice is to plan sample size *before* collecting data and to record the plan in your notebook. (Here, because this is a teaching template, we do the calculation alongside the data.) For a paired test, the relevant effect size is $d_z$, and `pwr.t.test()` handles it with `type = "paired"`.

```{r}
#| label: power-plan
# A priori: how many PAIRS for 80% power to detect a large effect (dz = 0.8)?
plan <- pwr.t.test(d = 0.8, sig.level = alpha, power = target_power,
                   type = "paired", alternative = "two.sided")

# Sensitivity: with the pairs we actually have, what is the smallest effect
# we could detect with 80% power?
sens <- pwr.t.test(n = n_pairs, sig.level = alpha, power = target_power,
                   type = "paired", alternative = "two.sided")

# Power to detect a large (dz = 0.8) and a medium (dz = 0.5) effect with our n
power_large  <- pwr.t.test(n = n_pairs, d = 0.8, sig.level = alpha, type = "paired")$power
power_medium <- pwr.t.test(n = n_pairs, d = 0.5, sig.level = alpha, type = "paired")$power
```

To detect a large effect ($d_z = 0.8$) with 80% power at $\alpha = .05$, we would need about **`r ceiling(plan$n)` pairs** (always round up). With `r n_pairs` pairs, the smallest effect we could detect with 80% power is **$d_z$ = `r fmt_num(sens$d)`**. The power to detect a true large effect is about **`r round(100 * power_large)`%**, and for a medium effect about **`r round(100 * power_medium)`%**. This study was therefore **underpowered**, even for a large effect (the conventional target is 80%), which we should keep in mind when interpreting the results.

### Why pairing can help (and when it does not) {#sec-pairing-power}

Is a paired design worth the trouble? The answer depends on how strongly the two measurements from the same animal are related. The standard deviation of the differences depends on the two conditions' standard deviations ($s_1$, $s_2$) and on their correlation $r$:

$$s_D^2 = s_1^2 + s_2^2 - 2\,r\,s_1 s_2 .$$

When animals that travel far under the vehicle also travel far under the drug ($r$ is large and positive), the subtraction **cancels** the stable animal-to-animal differences and $s_D$ shrinks. That makes $d_z$ larger for the same raw effect, and power rises. If $r$ is near 0, there is nothing to cancel and the paired design gains nothing (and loses a little, because it has $n - 1$ rather than about $2n - 2$ degrees of freedom). @tbl-pairing shows this for a medium effect measured in the usual two-sample way ($d_{av} = 0.5$, the mean difference divided by the typical SD of a single condition).

```{r}
#| label: tbl-pairing
#| tbl-cap: "How the correlation between a mouse's two measurements changes power for the same raw effect (d = 0.5 in single-condition SD units)"
pairing <- tibble(r = c(0, 0.3, 0.6, 0.9)) |>
  mutate(
    d_z = 0.5 / sqrt(2 * (1 - r)),
    power_paired = map_dbl(d_z, ~ pwr.t.test(n = n_pairs, d = .x, sig.level = alpha,
                                             type = "paired")$power),
    pairs_needed = map_dbl(d_z, ~ ceiling(pwr.t.test(d = .x, power = target_power,
                                                     sig.level = alpha,
                                                     type = "paired")$n))
  )

pairing |>
  transmute(`Correlation between conditions (r)` = fmt_num(r, 1),
            `Implied dz`                         = fmt_num(d_z),
            `Power with our number of pairs`     = scales::percent(power_paired, accuracy = 1),
            `Pairs needed for 80% power`         = pairs_needed) |>
  pub_table(align = "cccc")
```

For comparison, a design that tests **different** mice in each condition would need about `r ceiling(pwr.t.test(d = 0.5, power = target_power, sig.level = alpha, type = "two.sample")$n)` mice *per group* (twice that many in total) to detect the same medium effect. A paired design is therefore economical when the correlation is high: with $r$ = .6 it needs about `r pairing$pairs_needed[pairing$r == 0.6]` mice (each tested twice), against `r 2 * ceiling(pwr.t.test(d = 0.5, power = target_power, sig.level = alpha, type = "two.sample")$n)` mice for the independent-groups design. With $r$ near 0, pairing saves little: it needs about `r pairing$pairs_needed[pairing$r == 0]` mice, each tested twice.

# Results and Conclusions {#sec-results}

## Exploratory data analysis {#sec-eda}

Before running any test, we describe and plot the data. @tbl-desc gives the descriptive statistics for each condition (using only the `r n_pairs` complete pairs) and for the difference scores.

```{r}
#| label: tbl-desc
#| tbl-cap: "Descriptive statistics for distance traveled (m), by condition, and for the within-mouse difference (Drug − Vehicle)"
desc_long <- pairs |>
  select(mouse_id, Vehicle, Drug, Difference = diff) |>
  pivot_longer(-mouse_id, names_to = "Measure", values_to = "value") |>
  mutate(Measure = factor(Measure, levels = c("Vehicle", "Drug", "Difference")))

desc <- desc_long |>
  group_by(Measure) |>
  summarize(
    n      = n(),
    M      = mean(value),
    SD     = sd(value),
    SE     = SD / sqrt(n),
    Median = median(value),
    Min    = min(value),
    Max    = max(value),
    .groups = "drop"
  )

desc |> pub_table(digits = 2, align = "lcccccc")
```

```{r}
#| label: desc-objects
veh  <- desc |> filter(Measure == "Vehicle")
drg  <- desc |> filter(Measure == "Drug")
dif  <- desc |> filter(Measure == "Difference")
r_obs <- cor(pairs$Vehicle, pairs$Drug)
n_up  <- sum(pairs$diff > 0)
```

After the vehicle, mice traveled a mean of `r fmt_num(veh$M)` m (*SD* = `r fmt_num(veh$SD)`); after the drug, `r fmt_num(drg$M)` m (*SD* = `r fmt_num(drg$SD)`). The mean within-mouse difference was `r fmt_num(dif$M)` m (*SD* = `r fmt_num(dif$SD)`), and `r n_up` of the `r n_pairs` mice traveled farther after the drug. Notice the large spread **between** mice in both conditions, and that the two measurements from the same mouse are positively correlated ($r$ = `r fmt_apa(r_obs)`). That correlation is what the paired design takes advantage of (@sec-pairing-power).

```{r}
#| label: fig-eda
#| fig-cap: "Paired structure of the data. (A) Each point is one mouse; points above the diagonal line (no change) traveled farther after the drug than after the vehicle. (B) Each mouse's change in distance (Drug − Vehicle), sorted from largest to smallest; blue bars are increases and red-orange bars are decreases."
#| fig-alt: "Panel A is a scatter plot of distance after the drug against distance after the vehicle with a diagonal line of no change; most points lie above the line and the points rise together, showing a positive correlation. Panel B is a bar chart of each mouse's change in distance sorted from largest to smallest; most bars are positive and a few are negative."
#| fig-width: 7
#| fig-height: 3.8
lims <- range(c(pairs$Vehicle, pairs$Drug)) + c(-2, 2)

p_scatter <- ggplot(pairs, aes(Vehicle, Drug)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(size = 2.5, color = pal["blue"]) +
  coord_equal(xlim = lims, ylim = lims) +
  labs(x = "Distance after vehicle (m)", y = "Distance after drug (m)")

p_bars <- pairs |>
  mutate(direction = if_else(diff >= 0, "Increase", "Decrease"),
         mouse_id  = fct_reorder(mouse_id, diff, .desc = TRUE)) |>
  ggplot(aes(mouse_id, diff, fill = direction)) +
  geom_col() +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values = c(Increase = unname(pal["blue"]),
                               Decrease = unname(pal["vermillion"]))) +
  labs(x = "Mouse (sorted)", y = "Change in distance (m)", fill = NULL) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

p_scatter + p_bars + plot_annotation(tag_levels = "A")
```

## Checking the assumptions {#sec-assumptions}

For a paired t-test, the normality and outlier checks are made on the **difference scores**.

```{r}
#| label: assumptions
shapiro_diff <- tidy(shapiro.test(pairs$diff))

outliers <- pairs |> identify_outliers(diff)
```

```{r}
#| label: fig-assumptions
#| fig-cap: "Checking normality of the within-mouse differences. (A) Histogram. (B) Normal Q-Q plot; points close to the line indicate approximate normality."
#| fig-alt: "Panel A is a histogram of the within-mouse differences in distance, centered above zero, with a longer tail to the left (one mouse traveled about 13 meters less after the drug). Panel B is a normal Q-Q plot in which most points follow the line, but the lowest point lies below it."
#| fig-width: 7
#| fig-height: 3.4
p_h <- ggplot(pairs, aes(diff)) +
  geom_histogram(binwidth = 4, fill = pal["sky"], color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Drug − Vehicle (m)", y = "Number of mice")

p_q <- ggplot(pairs, aes(sample = diff)) +
  stat_qq(color = pal["blue"]) +
  stat_qq_line(color = "black") +
  labs(x = "Theoretical quantiles", y = "Sample quantiles")

p_h + p_q + plot_annotation(tag_levels = "A")
```

- **Normality.** The Shapiro–Wilk test `r if (shapiro_diff$p.value < alpha) "indicated" else "did not indicate"` a departure from normality in the differences (*W* = `r fmt_apa(shapiro_diff$statistic, 3)`, `r fmt_p(shapiro_diff$p.value, md = TRUE)`), and the Q-Q plot (@fig-assumptions) is roughly linear, apart from the one mouse that traveled much less far after the drug (the lowest point). With only `r n_pairs` pairs the histogram is rough, so the Q-Q plot is more informative.
- **Outliers.** `r nrow(outliers)` `r if (nrow(outliers) == 1) "difference score was" else "difference scores were"` flagged as outliers (more than 1.5 × IQR beyond the quartiles).
- **Independence.** Each pair comes from a different mouse. We assume the mice did not influence each other (for example, by sharing a cage with a stimulated cage-mate). The two sessions of one mouse are *supposed* to be related; the paired test accounts for that.

::: {.callout-warning}
## A non-significant assumption test is not proof the assumption holds
With small samples, tests such as Shapiro–Wilk have little power to detect problems, and with large samples they flag trivial ones. Judge the plots first and use the test as supporting evidence.
:::

## Hypothesis test {#sec-test}

```{r}
#| label: ttest
# paired = TRUE: the i-th Drug value is matched with the i-th Vehicle value, so
# the two columns must be in the same mouse order (they are: one row per mouse).
# The difference is Drug - Vehicle.
tt <- t.test(pairs$Drug, pairs$Vehicle, paired = TRUE,
             alternative = "two.sided", conf.level = 1 - alpha)

tt_tidy <- tidy(tt)   # one-row data frame (broom package)

t_value   <- tt_tidy$statistic
t_df      <- tt_tidy$parameter
p_value   <- tt_tidy$p.value
mean_diff <- tt_tidy$estimate        # mean of the differences (Drug - Vehicle)
ci_low    <- tt_tidy$conf.low
ci_high   <- tt_tidy$conf.high

# Effect size dz = mean(D) / SD(D). This is the one-sample Cohen's d applied to
# the difference scores, with a confidence interval based on the noncentral t.
dz <- effectsize::cohens_d(pairs$diff)
gz <- effectsize::hedges_g(pairs$diff)   # small-sample-corrected version
```

```{r}
#| label: tbl-ttest
#| tbl-cap: "Paired-samples t-test comparing distance traveled after the drug and after the vehicle"
tt_tidy |>
  transmute(
    `Mean difference` = mean_diff,
    `95% CI of difference` = paste0("[", fmt_num(ci_low), ", ", fmt_num(ci_high), "]"),
    t  = t_value,
    df = t_df,
    p  = if_else(p_value < .001, "< .001", sub("^0", "", fmt_num(p_value, 3))),
    `Cohen's dz`   = fmt_num(dz$Cohens_d),
    `95% CI of dz` = paste0("[", fmt_num(dz$CI_low), ", ", fmt_num(dz$CI_high), "]")
  ) |>
  pub_table(digits = 2, align = "c")
```

**Decision rule.** If $p < \alpha$, we reject $H_0$. Here `r fmt_p(p_value, md = TRUE)`, which is `r if (p_value < alpha) "less" else "not less"` than $\alpha$ = `r fmt_apa(alpha)`, so we **`r if (p_value < alpha) "reject" else "fail to reject"` $H_0$**.

::: {.callout-important}
## What a p-value is (and is not)
The p-value is the probability of obtaining a mean difference at least this far from zero **if $H_0$ were true**. It is *not* the probability that $H_0$ is true, not the probability the result happened "by chance," and not a measure of how large or important the effect is.
:::

## Effect size and confidence interval {#sec-effect}

The p-value tells us whether a difference is *detectable*; the **effect size** tells us how *large* it is.

```{r}
#| label: effect-objects
# d_av: mean difference in units of the typical SD of a SINGLE condition. This is
# the version that is comparable to the d of an independent-groups study.
d_av <- mean_diff / sqrt((veh$SD^2 + drg$SD^2) / 2)
```

- **Raw difference.** Mice traveled **`r fmt_num(abs(mean_diff))` m** `r if (mean_diff > 0) "farther" else "less far"` after the drug than after the vehicle (95% CI [`r fmt_num(ci_low)`, `r fmt_num(ci_high)`]). Relative to the vehicle mean, this is a change of about **`r round(100 * abs(mean_diff) / veh$M)`%**.
- **Standardized difference.** Cohen's $d_z$ was **`r fmt_num(dz$Cohens_d)`** (95% CI [`r fmt_num(dz$CI_low)`, `r fmt_num(dz$CI_high)`]). (Hedges' $g_z$, a small-sample correction, was `r fmt_num(gz$Hedges_g)`.)

The confidence interval for the difference `r if (ci_low > 0 || ci_high < 0) "does not include" else "includes"` 0, which agrees with the test decision. The interval for $d_z$ is wide (`r fmt_num(dz$CI_low)` to `r fmt_num(dz$CI_high)`): with only `r n_pairs` pairs, the true effect size is estimated imprecisely.

::: {.callout-note}
## Two kinds of "d" for paired data
$d_z$ divides the mean difference by the SD of the **differences**. That SD is small when the two measurements are strongly correlated, so $d_z$ grows with the correlation even when the raw effect is unchanged. Here $d_z$ = `r fmt_num(dz$Cohens_d)`, whereas the **$d_{av}$** that divides by the typical SD of a single condition is `r fmt_num(d_av)`. (They happen to be close here partly because the two conditions have different spreads. With equal SDs and a correlation above .5, as here, $d_z$ would be larger than $d_{av}$: in general $d_z = d_{av}/\sqrt{2(1-r)}$.) Only $d_{av}$ is comparable to the *d* from a two-sample study of the same drug. Cohen's small/medium/large labels were defined for between-group *d*, so use them for $d_z$ only as a rough guide. Whichever you report, **say which one it is**, and always report the raw difference in meters with its confidence interval, which is the most interpretable summary.
:::

```{r}
#| label: fig-main
#| fig-cap: "Distance traveled in the open field after the vehicle and after the drug. (A) Each gray line connects the two sessions of one mouse; blue and red-orange points show the individual values, and black markers show each condition's mean with its 95% confidence interval. (B) The estimated mean within-mouse difference (Drug − Vehicle) with its 95% confidence interval; the dashed line marks no difference."
#| fig-alt: "Panel A shows each mouse's distance after the vehicle and after the drug joined by a gray line; most lines slope upward, though a few slope downward, and black markers show a higher mean for the drug. Panel B shows the mean difference of about 5 meters with a confidence interval whose lower end is just above the dashed zero line."
#| fig-width: 7
#| fig-height: 4.5
cond_ci <- pairs |>
  select(mouse_id, Vehicle, Drug) |>
  pivot_longer(c(Vehicle, Drug), names_to = "treatment", values_to = "distance") |>
  mutate(treatment = factor(treatment, levels = c("Vehicle", "Drug")))

cond_sum <- cond_ci |>
  group_by(treatment) |>
  summarize(m = mean(distance), se = sd(distance) / sqrt(n()), n = n(), .groups = "drop") |>
  mutate(lo = m - qt(0.975, n - 1) * se,
         hi = m + qt(0.975, n - 1) * se)

fig_a <- ggplot(cond_ci, aes(treatment, distance)) +
  geom_line(aes(group = mouse_id), color = "gray60") +
  geom_point(aes(color = treatment), size = 2.4) +
  geom_pointrange(data = cond_sum, aes(y = m, ymin = lo, ymax = hi),
                  position = position_nudge(x = 0.2), color = "black",
                  shape = 16, size = 0.5, linewidth = 0.8) +
  scale_color_manual(values = unname(pal[c("blue", "vermillion")])) +
  labs(x = NULL, y = "Distance traveled (m)") +
  theme(legend.position = "none")

diff_df <- tibble(m = mean_diff, lo = ci_low, hi = ci_high)

fig_b <- ggplot(diff_df, aes(x = "Drug - Vehicle", y = m)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30") +
  geom_pointrange(aes(ymin = lo, ymax = hi), color = pal["green"],
                  size = 0.7, linewidth = 1) +
  labs(x = NULL, y = "Mean difference (m)") +
  theme(axis.ticks.x = element_blank())

fig_a + fig_b + plot_layout(widths = c(2, 1)) + plot_annotation(tag_levels = "A")
```

The black confidence intervals in panel A are for each condition's *mean*, and they overlap. Do not use that overlap to judge the paired test: the question is about the **differences**, which panel B shows, and the interval for the difference is what answers it. This is exactly the information that pairing preserves.

## What did pairing buy us? {#sec-pairing-result}

To see what the paired design contributed, we can analyze the same numbers the wrong way, as if the `r n_pairs` Drug values and the `r n_pairs` Vehicle values came from different mice.

```{r}
#| label: ignore-pairing
welch <- tidy(t.test(pairs$Drug, pairs$Vehicle, paired = FALSE))   # Welch, ignores pairing

# The SD of the differences can be rebuilt from the two SDs and their correlation
sd_d_check <- sqrt(veh$SD^2 + drg$SD^2 - 2 * r_obs * veh$SD * drg$SD)
```

```{r}
#| label: tbl-ignore
#| tbl-cap: "The same data analyzed as paired and (incorrectly) as two independent groups"
tibble(
  Analysis = c("Paired t-test (correct for this design)",
               "Welch two-sample t-test (ignores the pairing)"),
  `Mean difference` = c(mean_diff, welch$estimate),
  `95% CI` = c(paste0("[", fmt_num(ci_low), ", ", fmt_num(ci_high), "]"),
               paste0("[", fmt_num(welch$conf.low), ", ", fmt_num(welch$conf.high), "]")),
  t  = c(t_value, welch$statistic),
  df = c(fmt_num(t_df, 0), fmt_num(welch$parameter, 1)),
  p  = c(fmt_p_cell(p_value), fmt_p_cell(welch$p.value))
) |>
  pub_table(digits = 2, align = "lccccc")
```

The two analyses estimate the **same mean difference** (`r fmt_num(mean_diff)` m), but the paired analysis gives a narrower confidence interval and a smaller *p*-value (`r fmt_p(p_value, md = TRUE)` vs. `r fmt_p(welch$p.value, md = TRUE)`). The reason is the correlation between a mouse's two measurements: rebuilding $s_D$ from the two condition SDs and $r$ = `r fmt_apa(r_obs)` gives `r fmt_num(sd_d_check)`, matching the SD of the differences (`r fmt_num(dif$SD)`), which is much smaller than the spread of either condition.

::: {.callout-warning}
## Choose the test from the design, not from the p-value
The test must match **how the data were collected**. If each mouse was measured twice, analyze the pairs. If different mice were in each group, a paired test is *wrong* (there is nothing to pair), and it can give misleading results. Never try both and report the one you like better.
:::

## Order effects in a crossover design {#sec-order}

A within-subject design has its own threat: the second session might differ from the first for reasons unrelated to the drug, such as **habituation** to the arena (the second visit is less novel, so mice explore less), **practice**, or **carryover** of the first treatment. If every mouse got the vehicle first and the drug second, an order effect would be completely tangled up with the drug effect. **Counterbalancing**, giving half the mice each order, spreads the order effect evenly across both treatments, so it cancels in the average difference. @tbl-order checks this in our data.

```{r}
#| label: tbl-order
#| tbl-cap: "Mean within-mouse difference (Drug − Vehicle) by treatment sequence"
order_tbl <- pairs |>
  group_by(Sequence = sequence) |>
  summarize(Pairs = n(), `Mean difference` = mean(diff), `SD of differences` = sd(diff),
            .groups = "drop")

order_tbl |> pub_table(digits = 2, align = "lccc")
```

```{r}
#| label: fig-order
#| fig-cap: "Within-mouse difference (Drug − Vehicle) for mice that received the drug first or the vehicle first. Each point is one mouse; diamonds are group means; the dashed line marks no difference."
#| fig-alt: "A dot plot of within-mouse differences for the two treatment sequences. The differences for the drug-first mice sit higher than those for the vehicle-first mice, and most points in both groups are above the dashed zero line."
#| fig-width: 5
#| fig-height: 3.6
ggplot(pairs, aes(sequence, diff)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30") +
  geom_jitter(width = 0.08, height = 0, size = 2.4, color = pal["blue"]) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3.5,
               fill = "white", color = "black") +
  labs(x = "Sequence", y = "Difference in distance, Drug − Vehicle (m)")
```

If there is a session effect (say, session 2 is lower than session 1), the *drug-first* mice show the drug effect **plus** the session effect (because the vehicle session happened second and was lowered), while the *vehicle-first* mice show the drug effect **minus** it. Halving the gap between the two sequence means estimates the size of the session effect (assuming there is no carryover; with carryover, the gap also reflects it): here (`r fmt_num(order_tbl$"Mean difference"[order_tbl$Sequence == "Drug first"])` − `r fmt_num(order_tbl$"Mean difference"[order_tbl$Sequence == "Vehicle first"])`) / 2 = `r fmt_num((order_tbl$"Mean difference"[order_tbl$Sequence == "Drug first"] - order_tbl$"Mean difference"[order_tbl$Sequence == "Vehicle first"]) / 2)` m, consistent with mice being a little less active in the second session. The session effect also adds to the spread of the differences, which is one reason the SDs in @tbl-order are large. The cancellation in the overall average is exact only if the two sequences have the same number of animals. Because of the lost recording we have `r paste(order_tbl$Pairs, collapse = " and ")` pairs, so the cancellation is close but not perfect, and the sequence means are only a rough guide with so few mice.

::: {.callout-note}
## Designing a within-subject study
Counterbalance the order, randomize which mice get which order, allow a washout long enough for the first treatment to wear off, and keep conditions the same across sessions. Analyses that model sequence and session formally are possible (a "crossover analysis") but go beyond this course; discuss them with your instructor if your design has more than two treatments or sessions.
:::

## Power and sensitivity {#sec-power}

The power curves in @fig-power show how power grows with the number of pairs for small, medium, and large true effects.

```{r}
#| label: fig-power
#| fig-cap: "Statistical power of a two-tailed paired t-test (α = .05) as a function of the number of pairs, for small, medium, and large true effects (dz). The dashed horizontal line marks 80% power."
#| fig-alt: "Line graph of statistical power against number of pairs for small, medium, and large effects. Power rises with sample size, fastest for the large effect, which reaches the 80 percent reference line at about 15 pairs. The medium effect needs about 34 pairs, and the small effect stays well below 80 percent."
power_curve <- expand_grid(n = 4:60, d = c(0.2, 0.5, 0.8)) |>
  mutate(
    power   = map2_dbl(n, d, ~ pwr.t.test(n = .x, d = .y, sig.level = alpha,
                                          type = "paired")$power),
    d_label = factor(d, labels = c("Small (dz = 0.2)", "Medium (dz = 0.5)",
                                   "Large (dz = 0.8)"))
  )

ggplot(power_curve, aes(n, power, color = d_label, linetype = d_label)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = target_power, linetype = "dashed") +
  scale_color_manual(values = unname(pal[c("sky", "blue", "vermillion")])) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(x = "Number of pairs (mice)", y = "Power",
       color = "True effect size", linetype = "True effect size")
```

With `r n_pairs` pairs, the power to detect a true *large* effect ($d_z = 0.8$) was about **`r round(100 * power_large)`%**, so we would **miss** a real effect of that size (a Type II error) in roughly `r round(100 * (1 - power_large))`% of experiments like this one.

::: {.callout-caution}
## Significant results from small studies tend to exaggerate effects
When power is low, the studies that *do* reach significance are, on average, the ones whose samples happened to show an unusually large difference. A significant result from an underpowered study therefore tends to **overestimate** the true effect (sometimes called the "winner's curse"). That is one reason to rely on the confidence interval, not just the point estimate, and to look for replication.
:::

::: {.callout-caution}
## Avoid "observed power"
Computing power from the effect size you just observed adds no information beyond the p-value (the two are mathematically linked) and is widely discouraged. Use the **sensitivity analysis** above (what could we detect?) and the **confidence interval** (what effects are compatible with our data?) instead.
:::

## Type I and Type II errors in action {#sec-sim}

Error rates are long-run properties: they describe what would happen if we repeated the experiment many times. We can see this directly by **simulating** 10,000 experiments with `r n_pairs` pairs each, in two imaginary worlds where we know the truth:

1. **$H_0$ is true.** The mean difference is exactly zero.
2. **$H_1$ is true.** The mean difference is 0.8 standard deviations of the differences ($d_z = 0.8$, a large effect).

In each simulated experiment we draw `r n_pairs` difference scores from a normal distribution and run a paired t-test (which is a one-sample t-test on the differences).

```{r}
#| label: simulate
simulate_paired_p <- function(n, dz, n_sims = 10000) {
  map_dbl(seq_len(n_sims), function(i) {
    d <- rnorm(n, mean = dz, sd = 1)   # difference scores, in SD units
    t.test(d)$p.value                  # same as a paired t-test
  })
}

set.seed(123)   # makes the simulation reproducible
p_h0 <- simulate_paired_p(n_pairs, dz = 0)
p_h1 <- simulate_paired_p(n_pairs, dz = 0.8)

sim_summary <- tibble(
  World = c("H0 is true (no real difference)", "H1 is true (dz = 0.8)"),
  `Experiments with p < .05` = c(mean(p_h0 < alpha), mean(p_h1 < alpha)),
  `What this rate is` = c("Type I error rate", "Power")
)
```

```{r}
#| label: tbl-sim
#| tbl-cap: "Results of 10,000 simulated experiments (each with the same number of pairs as the analyzed sample)"
sim_summary |>
  mutate(`Experiments with p < .05` = scales::percent(`Experiments with p < .05`,
                                                      accuracy = 0.1)) |>
  pub_table(align = "lcl")
```

```{r}
#| label: fig-sim
#| fig-cap: "Distribution of p-values across 10,000 simulated experiments. When H0 is true (left), p-values are spread evenly between 0 and 1, so about 5% fall below .05 (Type I errors). When a large real effect exists (right), p-values pile up near 0, but those to the right of the dashed line (p ≥ .05) are Type II errors."
#| fig-alt: "Two histograms of p-values from 10,000 simulated experiments. When the null hypothesis is true, p-values are spread evenly from 0 to 1. When a large effect exists, p-values are concentrated near 0, but a substantial number still fall above .05."
#| fig-width: 7
#| fig-height: 3.5
bind_rows(
  tibble(p = p_h0, scenario = "H0 true (no real difference)"),
  tibble(p = p_h1, scenario = "H1 true (dz = 0.8)")
) |>
  ggplot(aes(p)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = pal["sky"], color = "white") +
  geom_vline(xintercept = alpha, linetype = "dashed", color = pal["vermillion"]) +
  facet_wrap(~ scenario) +
  labs(x = "p-value", y = "Number of simulated experiments")
```

What @tbl-sim and @fig-sim teach us:

- When $H_0$ is true, the paired t-test falsely rejects it in about **`r scales::percent(mean(p_h0 < alpha), accuracy = 0.1)`** of experiments, close to our chosen $\alpha$ of 5%. **The Type I error rate is something we set.**
- When a large effect really exists, the test detects it only about **`r scales::percent(mean(p_h1 < alpha), accuracy = 0.1)`** of the time (its power), so it commits a Type II error in about **`r scales::percent(mean(p_h1 >= alpha), accuracy = 0.1)`**. **The Type II error rate depends on the number of pairs, the effect size, and variability.**
- Simulated rates have a little random error of their own: with 10,000 experiments, a rate near 5% is accurate to within about ±0.4 percentage points.
- A single "significant" result does not tell you which world you are in. That is why effect sizes, confidence intervals, and replication matter.

## Robustness check

If we are unsure about the normality of the differences, we can confirm the conclusion with a test that does not assume it. The **Wilcoxon signed-rank test** ranks the sizes of the differences and asks whether the positive and negative ones balance out.

```{r}
#| label: wilcoxon
wilcox <- wilcox.test(pairs$Drug, pairs$Vehicle, paired = TRUE) |>
  tidy()
```

The Wilcoxon signed-rank test `r if ((wilcox$p.value < alpha) == (p_value < alpha)) "led to the same conclusion as" else "led to a different conclusion from"` the paired t-test, *V* = `r wilcox$statistic`, `r fmt_p(wilcox$p.value, md = TRUE)`. (Pairs with a difference of exactly zero are dropped by this test. R uses an exact calculation when there are no ties among the sizes of the differences, and a normal approximation otherwise.) `r if ((wilcox$p.value < alpha) != (p_value < alpha)) "The two p-values fall on opposite sides of .05, but they are close, so this is not a meaningful disagreement: the evidence is borderline by either method, and that is the honest message to report." else ""`

## Results write-up {#sec-writeup}

Below is how the result would be written in the Results section of a report. The numbers are inserted directly from the analysis (using inline R code), so they cannot be mistyped and they update automatically if the data change.

::: {.callout-note icon=false appearance="simple"}
### Results

Of the `r n_recorded` mice tested, the video tracking for one mouse's `r tolower(as.character(lost_row$treatment))` session failed, so that mouse was excluded from the paired analysis, leaving `r n_pairs` complete pairs. Mice traveled `r if (p_value < alpha) "significantly" else "not significantly"` `r if (mean_diff > 0) "farther" else "less far"` after the drug (*M* = `r fmt_num(drg$M)` m, *SD* = `r fmt_num(drg$SD)`) than after the vehicle (*M* = `r fmt_num(veh$M)` m, *SD* = `r fmt_num(veh$SD)`), paired *t*(`r t_df`) = `r fmt_num(t_value)`, `r fmt_p(p_value, md = TRUE)`. The mean within-mouse difference was `r fmt_num(mean_diff)` m, 95% CI [`r fmt_num(ci_low)`, `r fmt_num(ci_high)`], Cohen's *d*~z~ = `r fmt_num(dz$Cohens_d)`, 95% CI [`r fmt_num(dz$CI_low)`, `r fmt_num(dz$CI_high)`] (*d*~av~ = `r fmt_num(d_av)`). Individual results are shown in @fig-main. Treatment order was counterbalanced (`r paste(order_tbl$Pairs, collapse = " and ")` complete pairs in the two sequences). A Wilcoxon signed-rank test, which does not assume normal differences, `r if ((wilcox$p.value < alpha) == (p_value < alpha)) "gave the same conclusion" else "gave a different conclusion"` (`r fmt_p(wilcox$p.value, md = TRUE)`).
:::

## Conclusions and limitations {#sec-conclusions}

**Conclusion.** `r if (p_value < alpha) "The data provide evidence that mice traveled farther in the open field after the stimulant than after the vehicle, consistent with increased locomotor activity." else "The data do not provide sufficient evidence of a difference in distance traveled after the drug and after the vehicle. This does not show that the drug has no effect; the study may have been too small to detect a modest one."` `r if (p_value < alpha) "The estimated increase is imprecise, though: the confidence interval is wide and its lower end is close to zero, so the study cannot say how large the effect is." else ""`

::: {.callout-warning}
## Re-read the interpretation if your data change
The numbers (and any wording tied to them with inline R code) update automatically, but the *scientific interpretation* does not. Neither do hand-written sentences elsewhere in this notebook that describe a particular figure or table, or the figure descriptions (alt text). If you change the data or the analysis, re-read the conclusions, the limitations, and those descriptions, and revise them in your own words.
:::

**Limitations.**

- **Sample size.** With `r n_pairs` pairs, power for a large effect was only about `r round(100 * power_large)`%, and the confidence intervals for the effect are wide. Because the study is underpowered, the observed effect size may be an overestimate.
- **Order and carryover.** Counterbalancing protects the average difference from a session effect, but only approximately here (the sequences are unequal after the loss of one mouse), and it does not protect against *carryover* if the drug had not fully worn off after a week.
- **Missing data.** One mouse was lost because of a recording failure. A complete-case analysis is reasonable when the failure is unrelated to the outcome, which is plausible for an equipment problem.
- **One dose, one measure.** We tested a single dose and a single behavioral measure, so we cannot say how the effect depends on dose or whether other aspects of behavior change.
- **Sex.** Both sexes were tested, but `r n_pairs` pairs are far too few to say whether the drug works differently in females and males.
- **Simulated data.** These data were generated by a computer to illustrate the methods; they carry no real biological information.

# Special Topic: Many Measurements From One Animal {#sec-nested}

Neuroscience data often have a **nested** (clustered) structure: several neurons per animal, several brain slices per animal, several trials per session, or several pups per litter. The observations inside a cluster are **related**, because they share the same animal, genes, housing, and preparation. Treating them as independent is one of the most common statistical mistakes in the field, sometimes called **pseudoreplication**. This section shows what goes wrong and how to avoid it.

## The data {#sec-nested-data}

Suppose that in a second experiment, brain slices from **8 mice** were used to record neurons before and after bath application of the same drug. Several neurons were recorded from each mouse (3 to 6), and each neuron's **firing rate (Hz)** was measured at **baseline** and again **after the drug**. The data are in `Data/slice_recordings.csv`, in tidy long format (one row per neuron per condition).

```{r}
#| label: tbl-dictionary-slice
#| tbl-cap: "Variables in the slice recording file"
tribble(
  ~Variable,     ~Type,        ~Description,
  "mouse_id",    "Text",       "Unique animal identifier (S01–S08)",
  "neuron_id",   "Text",       "Unique neuron identifier (mouse ID plus neuron number)",
  "condition",   "Category",   "Baseline (before drug) or Drug (after bath application)",
  "firing_hz",   "Number",     "Firing rate in spikes per second (Hz)"
) |>
  pub_table(align = "lll")
```

```{r}
#| label: slice-import
slice_raw <- read_csv(here("Data", "slice_recordings.csv"), show_col_types = FALSE)

slice_wide <- slice_raw |>
  pivot_wider(names_from = condition, values_from = firing_hz) |>
  mutate(change = Drug - Baseline)       # Drug minus Baseline, one per neuron

stopifnot(!anyNA(slice_wide))            # this file has no missing values

n_neurons    <- nrow(slice_wide)
n_slice_mice <- n_distinct(slice_wide$mouse_id)

# One number per MOUSE: the average change across that mouse's neurons
mouse_means <- slice_wide |>
  group_by(mouse_id) |>
  summarize(neurons = n(), change = mean(change), .groups = "drop")
```

There are **`r n_neurons` neurons** but only **`r n_slice_mice` mice**. The question is: *how many independent observations do we have?*

```{r}
#| label: tbl-mouse-means
#| tbl-cap: "Neurons recorded and mean change in firing rate (Drug − Baseline) for each mouse"
mouse_means |>
  rename(Mouse = mouse_id, `Neurons recorded` = neurons, `Mean change (Hz)` = change) |>
  pub_table(digits = 2, align = "lcc")
```

```{r}
#| label: fig-nested
#| fig-cap: "Change in firing rate (Drug − Baseline) for every neuron, grouped by mouse. Each small point is one neuron; each black diamond is the mean of that mouse's neurons. The dashed line marks no change."
#| fig-alt: "A dot plot with one column per mouse showing each neuron's change in firing rate. Neurons from the same mouse cluster together, and the clusters sit at different heights; black diamonds mark each mouse's mean. Most clusters lie above the dashed zero line."
#| fig-width: 7
#| fig-height: 3.8
ggplot(slice_wide, aes(mouse_id, change)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30") +
  geom_jitter(width = 0.1, height = 0, size = 2, alpha = 0.8, color = pal["blue"]) +
  geom_point(data = mouse_means, shape = 23, size = 4, fill = "white", color = "black") +
  labs(x = "Mouse", y = "Change in firing rate (Hz)")
```

@fig-nested shows the problem. The neurons from one mouse tend to move together: the points for a mouse cluster around that mouse's own mean, and the means differ from mouse to mouse. A neuron is therefore partly "the same information again" as its neighbors from the same mouse.

## Two ways to analyze the same data

**Wrong: treat every neuron as an independent pair.** This gives a paired t-test with `r n_neurons` pairs. **Right: use the mouse as the unit.** Average the neuron-level changes within each mouse, so there is one difference per mouse, and run the paired (one-sample) t-test on those `r n_slice_mice` values. The animal is what was randomly sampled from the population and what the conclusion is about.

```{r}
#| label: nested-tests
naive   <- tidy(t.test(slice_wide$Drug, slice_wide$Baseline, paired = TRUE))   # n = neurons
correct <- tidy(t.test(mouse_means$change))                                    # n = mice

dz_mouse <- effectsize::cohens_d(mouse_means$change)
```

```{r}
#| label: tbl-nested
#| tbl-cap: "The same data analyzed with neurons (incorrect) and with mice (correct) as the unit of analysis"
tibble(
  Analysis = c("Neurons as units (pseudoreplication)", "Mice as units (average within mouse first)"),
  `Units (n)`         = c(naive$parameter + 1, correct$parameter + 1),
  `Mean change (Hz)`  = c(naive$estimate, correct$estimate),
  `95% CI` = c(paste0("[", fmt_num(naive$conf.low), ", ", fmt_num(naive$conf.high), "]"),
               paste0("[", fmt_num(correct$conf.low), ", ", fmt_num(correct$conf.high), "]")),
  t  = c(naive$statistic, correct$statistic),
  df = c(fmt_num(naive$parameter, 0), fmt_num(correct$parameter, 0)),
  p  = c(fmt_p_cell(naive$p.value), fmt_p_cell(correct$p.value))
) |>
  pub_table(digits = 2, align = "lcccccc")
```

The two analyses give similar estimates of the average change (`r fmt_num(naive$estimate)` Hz with neurons as units; `r fmt_num(correct$estimate)` Hz with mice as units; the small difference arises because mice contributed different numbers of neurons), but very different amounts of certainty. With neurons as units, the *p*-value is `r fmt_p(naive$p.value, md = TRUE)` and the confidence interval is narrow; with mice as units, it is `r fmt_p(correct$p.value, md = TRUE)` and the interval is `r fmt_num((correct$conf.high - correct$conf.low) / (naive$conf.high - naive$conf.low), 1)` times as wide. The neuron-level analysis pretends to have `r n_neurons` independent pieces of evidence when the study really tested the drug on `r n_slice_mice` animals.

```{r}
#| label: fig-nested-ci
#| fig-cap: "Estimated mean change in firing rate (Drug − Baseline) with 95% confidence intervals from the two analyses. The dashed line marks no change."
#| fig-alt: "Two points with error bars. The estimate with neurons as the unit has a narrow interval that stays well above the dashed zero line; the estimate with mice as the unit has a much wider interval that comes close to zero."
#| fig-width: 6
#| fig-height: 3
tibble(Analysis = factor(c("Neurons as units\n(incorrect)", "Mice as units\n(correct)"),
                         levels = c("Mice as units\n(correct)", "Neurons as units\n(incorrect)")),
       m  = c(naive$estimate, correct$estimate),
       lo = c(naive$conf.low, correct$conf.low),
       hi = c(naive$conf.high, correct$conf.high)) |>
  ggplot(aes(m, Analysis, color = Analysis)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray30") +
  geom_pointrange(aes(xmin = lo, xmax = hi), linewidth = 0.9, size = 0.6) +
  scale_color_manual(values = c(unname(pal["green"]), unname(pal["vermillion"]))) +
  labs(x = "Mean change in firing rate (Hz)", y = NULL) +
  theme(legend.position = "none")
```

## How much does clustering matter? Intraclass correlation and the design effect {#sec-icc}

The **intraclass correlation (ICC)** measures how similar observations from the same cluster are: it is the share of the total variation that is between mice rather than between neurons within a mouse. An ICC near 0 means neurons behave independently; an ICC near 1 means all neurons from a mouse are essentially copies of each other. The **design effect**, $1 + (\bar{m} - 1)\,\text{ICC}$, where $\bar{m}$ is the average number of neurons per mouse, says by what factor the variance of the estimated mean is larger than it would be with independent observations. Dividing the number of neurons by the design effect gives the **effective sample size**.

```{r}
#| label: icc
# Variance components for the neuron-level change, from a one-way ANOVA
# (simple "method of moments" estimates; mixed models do this more carefully)
aov_tab <- anova(lm(change ~ mouse_id, data = slice_wide))
MSB <- aov_tab["mouse_id", "Mean Sq"]       # mean square between mice
MSW <- aov_tab["Residuals", "Mean Sq"]      # mean square within mice
sizes <- as.integer(table(slice_wide$mouse_id))
n0 <- (n_neurons - sum(sizes^2) / n_neurons) / (n_slice_mice - 1)   # adjusted average size

sd_between <- sqrt(max((MSB - MSW) / n0, 0))   # SD between mouse means (true scale)
sd_within  <- sqrt(MSW)                        # SD between neurons within a mouse
icc        <- sd_between^2 / (sd_between^2 + sd_within^2)

m_bar <- n_neurons / n_slice_mice
deff  <- 1 + (m_bar - 1) * icc
n_eff <- n_neurons / deff
```

The estimated ICC was **`r fmt_apa(icc)`**: about `r round(100 * icc)`% of the variation in neuron-level change was between mice. With an average of `r fmt_num(m_bar)` neurons per mouse, the design effect is `r fmt_num(deff)` (an approximation, because mice contributed different numbers of neurons), so our `r n_neurons` neurons carry about as much information as **`r fmt_num(n_eff, 0)` independent neurons**, which is more than the `r n_slice_mice` mice but only about `r round(100 * n_eff / n_neurons)`% of the `r n_neurons` neurons. (With only `r n_slice_mice` mice the ICC itself is estimated imprecisely.)

## What does ignoring clustering do to error rates? {#sec-nested-sim}

To see the consequence, we simulate a world in which **the drug has no effect at all** ($H_0$ true), but the data have the same structure as ours: the same numbers of neurons per mouse and the variance components we just estimated. We then analyze each simulated experiment both ways and count how often each analysis (falsely) rejects $H_0$.

```{r}
#| label: nested-sim
simulate_nested <- function(sizes, sd_b, sd_w, n_sims = 5000) {
  id <- rep(seq_along(sizes), times = sizes)
  map_dfr(seq_len(n_sims), function(i) {
    u   <- rnorm(length(sizes), 0, sd_b)           # each mouse's own shift (no drug effect on average)
    chg <- rnorm(sum(sizes), mean = u[id], sd = sd_w)
    tibble(neurons = t.test(chg)$p.value,
           mice    = t.test(tapply(chg, id, mean))$p.value)
  })
}

set.seed(2468)
nested_sim <- simulate_nested(sizes, sd_between, sd_within)
```

```{r}
#| label: tbl-nested-sim
#| tbl-cap: "Type I error rates when the drug truly has no effect (5,000 simulated experiments with the same clustered structure as our data)"
tibble(
  Analysis = c("Neurons as units (incorrect)", "Mice as units (correct)"),
  `Experiments with p < .05` = scales::percent(c(mean(nested_sim$neurons < alpha),
                                                 mean(nested_sim$mice < alpha)),
                                               accuracy = 0.1)
) |>
  pub_table(align = "lc")
```

When there is truly no effect, the correct analysis falsely rejects $H_0$ in about **`r scales::percent(mean(nested_sim$mice < alpha), accuracy = 0.1)`** of experiments, close to the nominal 5%. Treating neurons as independent falsely rejects in about **`r scales::percent(mean(nested_sim$neurons < alpha), accuracy = 0.1)`**: a Type I error rate about `r fmt_num(mean(nested_sim$neurons < alpha) / alpha, 1)` times the 5% we think we have. The more neurons per mouse and the higher the ICC, the worse this gets. **Recording more neurons from the same animals does not substitute for testing more animals.**

## What to do in your own project {#sec-nested-advice}

- **Decide the unit of analysis before collecting data.** It is almost always the *independent* experimental unit: the animal (or the litter, or the culture batch). Write it in your analysis plan.
- **Simplest valid approach: average within animal first**, as above, then use the test that fits the design (paired, one-sample, two-sample, ANOVA) on those animal-level values. This is what we recommend in this course. It is valid even when animals have different numbers of neurons, though animals with few neurons give noisier means.
- **A more flexible approach:** a **mixed-effects (multilevel) model** keeps the neuron-level data and includes a random effect for mouse, for example `lme4::lmer(change ~ 1 + (1 | mouse_id), data = slice_wide)`. It handles unequal numbers of neurons per mouse and extra predictors, but with balanced data it gives results very close to averaging within mouse, and with only a handful of mice its estimates are shaky. Mixed models are the standard way to analyze nested data, but they need more statistical background and care than we can cover here. Consult your instructor before using one.
- **Always report both numbers**, for example "34 neurons from 8 mice", and make clear that the inferential statistics are based on the number of animals.
- **Don't pool across animals just to make the *n* look big.** The same issue arises with several slices per animal, several trials per subject, and several pups per litter.

::: {.callout-note icon=false appearance="simple"}
### Results (nested-data example)

Firing rates were recorded from `r n_neurons` neurons in slices from `r n_slice_mice` mice (3 to 6 neurons per mouse). Because neurons from the same mouse are not independent, each neuron's change in firing rate (Drug − Baseline) was averaged within mouse, and the `r n_slice_mice` mouse means were analyzed with a paired t-test. Firing increased by a mean of `r fmt_num(correct$estimate)` Hz after the drug, 95% CI [`r fmt_num(correct$conf.low)`, `r fmt_num(correct$conf.high)`], *t*(`r correct$parameter`) = `r fmt_num(correct$statistic)`, `r fmt_p(correct$p.value, md = TRUE)`, Cohen's *d*~z~ = `r fmt_num(dz_mouse$Cohens_d)`, 95% CI [`r fmt_num(dz_mouse$CI_low)`, `r fmt_num(dz_mouse$CI_high)`]. The intraclass correlation was `r fmt_apa(icc)`, so a neuron-level analysis would have overstated the evidence.
:::

# A Guide to Writing Up Results {#sec-guide}

A good results paragraph lets a reader understand and check what you did without seeing your code.

**Always report**

1. The **number of animals and the number of complete pairs** analyzed (and any exclusions or missing data, with the reason).
2. **Descriptive statistics for each condition**: the mean and standard deviation (*M*, *SD*).
3. The **name of the test** (paired-samples *t*-test), the **test statistic with degrees of freedom**, the **exact p-value**, and the **direction** of the change.
4. The **raw mean difference with a confidence interval**, and a **standardized effect size** with a confidence interval. For paired data, say whether it is *d*~z~ or *d*~av~.
5. For crossover designs, how the **order was counterbalanced**.
6. A reference to your **figure or table**. A good paired figure shows each subject's two values joined by a line.

**Style tips**

- Use past tense for what you did and found ("Mice traveled farther…").
- Write statistics in italics: *M*, *SD*, *t*, *p*, *d*. Do not put a zero before a p-value (*p* = .008, not 0.008).
- Report exact p-values rather than "p < .05"; use "*p* < .001" for very small ones.
- Say which condition was higher or lower. "The conditions differed" is not enough.
- Avoid "proves," "highly significant," or "trend toward significance." Say what you found and how large it was.
- "Significant" means only that *p* < α. Use the effect size to talk about importance.

**A fill-in-the-blank template**

```
Participants [did / did not] differ significantly between [Condition A]
(M = __, SD = __) and [Condition B] (M = __, SD = __) in [outcome], paired
t(__) = __, p = __, based on __ complete pairs. The mean difference
([A] − [B]) was __ [units], 95% CI [__, __] (Cohen's dz = __, 95% CI [__, __]),
a [small / medium / large] effect.
```

**If the result is not significant**, say so plainly and do not conclude that there is "no difference." Report the confidence interval (which differences are still compatible with the data?) and the sensitivity analysis (what size of effect could the study detect?). For example, in this study the smallest effect detectable with 80% power was $d_z$ = `r fmt_num(sens$d)`, so smaller (but possibly important) changes could not have been ruled out.

# Your Turn {#sec-your-turn}

Change one thing at a time, re-render, and describe in the notebook log (below) what changed and why.

1. **Ignore the pairing.** Look at @tbl-ignore. Rerun the analysis with `paired = FALSE`. How do the confidence interval and *p*-value change, and why? Then explain why using a paired test on two *independent* groups would be wrong.
2. **Missing data.** In the data preparation, replace the missing value with 0 instead of dropping the mouse. How do the means, the test, and the effect size change? Why is this the wrong way to handle a missing value? What if you had instead kept the mouse's one good session as an "extra" unpaired value?
3. **Sample size planning.** Use `pwr.t.test(type = "paired")` to find how many pairs you would need for 90% power to detect $d_z = 0.8$, and for 80% power to detect $d_z = 0.5$. How would a correlation of .8 between conditions change what you need? (Use @tbl-pairing as a guide.)
4. **One-tailed test.** Add `alternative = "greater"` to `t.test()`. What happens to the p-value? When would that be justified, and when would it be questionable?
5. **Order effects.** Suppose all 14 mice had received the vehicle first and the drug second. Compare this with @tbl-order: could you still separate the drug effect from the effect of session order? Why or why not?
6. **Outlier.** Change one mouse's drug value to an extreme number. What happens to the mean difference, the confidence interval, the *p*-value, and the Wilcoxon result? Which is more affected?
7. **Nested data.** Change the number of neurons per mouse in the simulation (in `sizes`, or by editing `01_simulate_data.R`) and the ICC. What happens to the Type I error rate of the neuron-level analysis? What if the ICC is close to 0?
8. **Your own data.** Replace the CSV in `Data/` with a file of your own (one row per subject per condition), update the file and variable names, and write your own problem statement, methods, and conclusions. If you have several measurements per animal, decide on your unit of analysis first.
9. **Interpretation.** Rewrite the Conclusion paragraph in your own words. What would you do next to follow up on this finding?

# Notebook Log {#sec-log}

Keep a running record of decisions so that you (and your instructor) can see how the analysis developed. Add a row whenever you change something.

```{r}
#| label: tbl-log
#| tbl-cap: "Analysis decision log (edit the rows below)"
tribble(
  ~Date,          ~`Decision or change`,                                        ~Reason,
  "YYYY-MM-DD",   "Chose a two-tailed paired t-test, α = .05",                   "Each mouse was tested under both conditions; the drug could change activity in either direction",
  "YYYY-MM-DD",   "Dropped mouse M07 from the paired analysis",                  "Tracking failed in one session, so the pair is incomplete; recorded as NA, not 0",
  "YYYY-MM-DD",   "Used mice (not neurons) as the unit in the slice example",    "Neurons from one mouse are not independent; averaged within mouse first",
  "",             "",                                                            ""
) |>
  pub_table(align = "lll")
```

# Reproducibility {#sec-repro}

The R and package versions used to create this notebook are listed below.

```{r}
#| label: session-info
sessionInfo()
```
